---
title: 'Evolving Event-driven Programs with SignalGP: <br/> Statistics'
output:
  pdf_document: default
  html_document: default
---

Here, we provide fully detailed statistical test results, including the R code used to generate them. Across experiments, we performed a Kruskal-Wallis rank sum test to determine if a set was significant, and if significant, we performed a post-hoc Dunn's test, applying a Bonferroni correction for multiple comparisons. See the paper for full context. 

Each post-hoc Dunn's test was performed using the FSA package.
```{r, echo=TRUE}
library(FSA)
```

```{r, echo=FALSE}
sig_thresh <- 0.05
```
We'll set our significance threshold to `r sig_thresh`.

## Changing Environment Problem

In this section, we give the statistical details for the Changing Environment Problem broken down by environment-size. 

First, we'll load the fitness data. 
```{r}
ce_ff_data <-  read.csv("../data/chg_env/mt_final_fitness.csv")
```

A note about how treatments are named within the data: treatment names describe the parameters and their values used when running the experiment. Parameters and their values are adjacent in the name, and parameter-value combinations are separated by underscores. For example, ED1_AS1_ENV2_TSK0 indicates that event-driven (ED) signals were enabled (1), active sensors (AS) were enabled (1), and there were two environments states. In other words, ED1_AS1_ENV2_TSK0 indicates the two-state environment combined treatment. The trailing TSK0 can be ignored. 

Okay, next we'll partition out different subsets of the data.
```{r}
# two-state environment 
# -- overall data --
ce_ff_data_2 <- ce_ff_data[grep("_ENV2_", ce_ff_data$treatment),]
ce_ff_data_2 <- ce_ff_data_2[ce_ff_data_2$analysis == "fdom",]
ce_ff_data_2$treatment <- relevel(ce_ff_data_2$treatment, ref="ED1_AS1_ENV2_TSK0")
# -- teasing apart combined treatment data --
ce_tff_data_2 <- ce_ff_data[grep("ED1_AS1_ENV2_", ce_ff_data$treatment),]
ce_tff_data_2$treatment <- relevel(ce_tff_data_2$treatment, ref="ED1_AS1_ENV2_TSK0")

# four-state environment
# -- overall data --
ce_ff_data_4 <- ce_ff_data[grep("_ENV4_", ce_ff_data$treatment),]
ce_ff_data_4 <- ce_ff_data_4[ce_ff_data_4$analysis == "fdom",]
ce_ff_data_4$treatment <- relevel(ce_ff_data_4$treatment, ref="ED1_AS1_ENV4_TSK0")
# -- teasing apart combined treatment data --
ce_tff_data_4 <- ce_ff_data[grep("ED1_AS1_ENV4_", ce_ff_data$treatment),]
ce_tff_data_4$treatment <- relevel(ce_tff_data_4$treatment, ref="ED1_AS1_ENV4_TSK0")

# eight-state environment
# -- overall data --
ce_ff_data_8 <- ce_ff_data[grep("_ENV8_", ce_ff_data$treatment),]
ce_ff_data_8 <- ce_ff_data_8[ce_ff_data_8$analysis == "fdom",]
ce_ff_data_8$treatment <- relevel(ce_ff_data_8$treatment, ref="ED1_AS1_ENV8_TSK0")
# -- teasing apart combined treatment data --
ce_tff_data_8 <- ce_ff_data[grep("ED1_AS1_ENV8_", ce_ff_data$treatment),]
ce_tff_data_8$treatment <- relevel(ce_tff_data_8$treatment, ref="ED1_AS1_ENV8_TSK0")

# sixteen-state environment
# -- overall data --
ce_ff_data_16 <- ce_ff_data[grep("_ENV16_", ce_ff_data$treatment),]
ce_ff_data_16 <- ce_ff_data_16[ce_ff_data_16$analysis == "fdom",]
ce_ff_data_16$treatment <- relevel(ce_ff_data_16$treatment, ref="ED1_AS1_ENV16_TSK0")
# -- teasing apart combined treatment data --
ce_tff_data_16 <- ce_ff_data[grep("ED1_AS1_ENV16_", ce_ff_data$treatment),]
ce_tff_data_16$treatment <- relevel(ce_tff_data_16$treatment, ref="ED1_AS1_ENV16_TSK0")
```

### Two-state Environment
#### Overall Results
```{r,echo=TRUE}
# Run the Kruskal-Wallis rank sum test
kw_fit_ce_2 <- kruskal.test(fitness ~ treatment, data=ce_ff_data_2)
kw_fit_ce_2
```
According to our Kruskal-Wallis rank sum test, at least one treatment is significantly different from the other treatments. Thus, we'll perform a post-hoc Dunn's test, applying a Bonferroni correction for multiple comparisons.

```{r, echo=TRUE}
# Run the post-hoc Dunn's test.
dt_ce_2 <- dunnTest(fitness~treatment, data=ce_ff_data_2, method="bonferroni")
dt_ce_2
```


#### Combined Treatment Re-evaluations
```{r,echo=TRUE}
# Run the Kruskal-Wallis rank sum test
kw_ce_comb_2 <- kruskal.test(fitness ~ analysis, data=ce_tff_data_2)
kw_ce_comb_2
```
According to our Kruskal-Wallis rank sum test, at least one treatment is significantly different from the other treatments. Thus, we'll perform a post-hoc Dunn's test, applying a Bonferroni correction for multiple comparisons.
```{r, echo=TRUE}
# Run the post-hoc Dunn's test.
dt_ce_comb_2 <- dunnTest(fitness~analysis, data=ce_tff_data_2, method="bonferroni")
dt_ce_comb_2
```


### Four-state Environment
#### Overall Results
```{r,echo=TRUE}
# Run the Kruskal-Wallis rank sum test
kw_fit_ce_4 <- kruskal.test(fitness ~ treatment, data=ce_ff_data_4)
kw_fit_ce_4
```
According to our Kruskal-Wallis rank sum test, at least one treatment is significantly different from the other treatments. Thus, we'll perform a post-hoc Dunn's test, applying a Bonferroni correction for multiple comparisons.

```{r, echo=TRUE}
# Run the post-hoc Dunn's test.
dt_ce_4 <- dunnTest(fitness~treatment, data=ce_ff_data_4, method="bonferroni")
dt_ce_4
```


#### Combined Treatment Re-evaluations
```{r,echo=TRUE}
# Run the Kruskal-Wallis rank sum test
kw_ce_comb_4 <- kruskal.test(fitness ~ analysis, data=ce_tff_data_4)
kw_ce_comb_4
```
According to our Kruskal-Wallis rank sum test, at least one treatment is significantly different from the other treatments. Thus, we'll perform a post-hoc Dunn's test, applying a Bonferroni correction for multiple comparisons.
```{r, echo=TRUE}
# Run the post-hoc Dunn's test.
dt_ce_comb_4 <- dunnTest(fitness~analysis, data=ce_tff_data_4, method="bonferroni")
dt_ce_comb_4
```


### Eight-state Environment
#### Overall Results
```{r,echo=TRUE}
# Run the Kruskal-Wallis rank sum test
kw_fit_ce_8 <- kruskal.test(fitness ~ treatment, data=ce_ff_data_8)
kw_fit_ce_8
```
According to our Kruskal-Wallis rank sum test, at least one treatment is significantly different from the other treatments. Thus, we'll perform a post-hoc Dunn's test, applying a Bonferroni correction for multiple comparisons.

```{r, echo=TRUE}
# Run the post-hoc Dunn's test.
dt_ce_8 <- dunnTest(fitness~treatment, data=ce_ff_data_8, method="bonferroni")
dt_ce_8
```


#### Combined Treatment Re-evaluations
```{r,echo=TRUE}
# Run the Kruskal-Wallis rank sum test
kw_ce_comb_8 <- kruskal.test(fitness ~ analysis, data=ce_tff_data_8)
kw_ce_comb_8
```
According to our Kruskal-Wallis rank sum test, at least one treatment is significantly different from the other treatments. Thus, we'll perform a post-hoc Dunn's test, applying a Bonferroni correction for multiple comparisons.
```{r, echo=TRUE}
# Run the post-hoc Dunn's test.
dt_ce_comb_8 <- dunnTest(fitness~analysis, data=ce_tff_data_8, method="bonferroni")
dt_ce_comb_8
```


### Sixteen-state Environment
#### Overall Results
```{r,echo=TRUE}
# Run the Kruskal-Wallis rank sum test
kw_fit_ce_16 <- kruskal.test(fitness ~ treatment, data=ce_ff_data_16)
kw_fit_ce_16
```
According to our Kruskal-Wallis rank sum test, at least one treatment is significantly different from the other treatments. Thus, we'll perform a post-hoc Dunn's test, applying a Bonferroni correction for multiple comparisons.

```{r, echo=TRUE}
# Run the post-hoc Dunn's test.
dt_ce_16 <- dunnTest(fitness~treatment, data=ce_ff_data_16, method="bonferroni")
dt_ce_16
```


#### Combined Treatment Re-evaluations
```{r,echo=TRUE}
# Run the Kruskal-Wallis rank sum test
kw_ce_comb_16 <- kruskal.test(fitness ~ analysis, data=ce_tff_data_16)
kw_ce_comb_16
```
According to our Kruskal-Wallis rank sum test, at least one treatment is significantly different from the other treatments. Thus, we'll perform a post-hoc Dunn's test, applying a Bonferroni correction for multiple comparisons.
```{r, echo=TRUE}
# Run the post-hoc Dunn's test.
dt_ce_comb_16 <- dunnTest(fitness~analysis, data=ce_tff_data_16, method="bonferroni")
dt_ce_comb_16
```

## Distributed Leader Election Problem
In this section, we give the statistical details for the distributed leader election problem. 

First, we'll load the fitness data. 
```{r}
election_data <- read.csv("../data/election/final_fitness.csv")
```

```{r,echo=TRUE}
# Run the Kruskal-Wallis rank sum test.
elec_kw <- kruskal.test(max_fitness ~ treatment, data=election_data)
elec_kw
```

```{r,echo=TRUE}
# Run the post-hoc Dunn's test.
elec_dt <- dunnTest(max_fitness~treatment, data=election_data, method="bonferroni")
elec_dt
```


